A free/open-source Kazakh-Tatar machine translation system

### 2013-04-21 paper/paper.tex
some scribbles on evaluation

%FIXME: describe new stuff
To get an idea of the kind of performance that could be expected from the system, we 
translated a simple story from Tatar to Bashkir and vice versa. The story may be found 
online,\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/branches/xupaixkar/rasskaz}}
and was used for pedagogical purposes in a recently workshop on MT
for the languages of Russia.

@@

Table~\ref{table:wer} presents the Word Error Rate, an edit metric based on the Levenshtein 
distance \citep{levenshtein/1966}. This measure was calculated once all the stems in the 
text had been added to the system, thus presents an upper bound on the current performance
of the transfer lexicon, and the disambiguation and transfer rules. The difference in 
the number of unknown words between translating Tatar$\rightarrow$Bashkir and vice versa
is because certain forms were not found due to lack of corresponding morphophonological rules.

We calculate the WER instead of other MT evaluation metrics such as BLEU as the WER is 
geared towards a particular task, that of measuring postedition effort. The translations 
of the story into Tatar and Bashkir were done in parallel to make them as close as possible,
so using BLEU would give an over-optimistic view of the quality.

@@

The majority of errors are currently due to mistakes and gaps in the morphophonology component; some minor problems still remain involving:
\begin{itemise}
  \item Combinations of case and possessive suffixes,
  \item Orthographical representations of phonology,
  \item Vowel harmony processing on clitics (e.g., \emph{да}/\emph{дә} `and') after unknown words.
\end{itemise}

@@

To our knowledge we have presented the first ever MT system between Tatar and Bashkir, and the first ever MT system involving Bashkir. The system is available as free/open-source software under the GNU GPL and the 
whole system may be downloaded from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/nursery/apertium-tt-ba}}

>>>>

To measure the performance of the translator we used the Word Error Rate metiric -- an edit metric based on the
Levenshtein distance \citep{levenshtein/1966}.

We had two small Kazakh corporas alongside with their postedited translations into Tatar to measure the WER.
The first one (2457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
on MT for the languages of Russia. In addition to postediting the translation, this corpus was run through the
morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
the system, and all the rules (Constraint Grammar,lexical selection and transfer) were based on this corpus.
Hence we can call it the development corpus. It presents an upper bound on the current performance
of the system. Table FIXME http://pastebin.com/TaaWxJzN presents the WER for it.

The second corpus -- a collection of only Radio Free Europe / Radio Liberty's
Kazakh-language service articles -- was used solely for evaluation. It had a similar size -- 2862 words.
The WER results for it can be seen in the table FIXME http://pastebin.com/7tzGCgMX.

Part of the corrections while postediting were due the lack of morphophonoligical rules in Tatar transducer,
and were not translation errors as such.

%The WER on the translation of the development corpus and its postedited version is twice as low.
%Which, considering the relatively low number of rules, indicates that a high quality translation can be achived
%by advancing these two modules.

%To get an idea of the kind of performance that could be expected from the system, we 
%translated a simple story from Tatar to Bashkir and vice versa. The story may be found 
%online,\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/branches/xupaixkar/rasskaz}}
%and was used for pedagogical purposes in a recently workshop on MT
%for the languages of Russia.

@@

%Table~\ref{table:wer} presents the Word Error Rate, an edit metric based on the Levenshtein 
%distance \citep{levenshtein/1966}. This measure was calculated once all the stems in the 
%text had been added to the system, thus presents an upper bound on the current performance
%of the transfer lexicon, and the disambiguation and transfer rules.

%We calculate the WER instead of other MT evaluation metrics such as BLEU as the WER is 
%geared towards a particular task, that of measuring postedition effort. The translations 
%of the story into Tatar and Bashkir were done in parallel to make them as close as possible,
%so using BLEU would give an over-optimistic view of the quality.

@@

%The majority of errors are currently due to mistakes and gaps in the morphophonology component; some minor problems still remain involving:
%\begin{itemise}
%  \item Combinations of case and possessive suffixes,
%  \item Orthographical representations of phonology,
%  \item Vowel harmony processing on clitics (e.g., \emph{да}/\emph{дә} `and') after unknown words.
%\end{itemise}

@@

To our knowledge we have presented the first ever MT system between Kazakh and Tatar. It has a near-to-production level
coverage, but relatively small number of rules

The system is available as free/open-source software under the GNU GPL and the whole system may be downloaded
from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
